---
title: "Phylogenetic Diversity - Communities"
author: "Student Name; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW
Taxonomic measures of $\alpha$- and $\beta$-diversity do not integrate evolutionary information. 
However, evolutionary information is important for a broad range of biodiversity issues including conservation, biogeography, and community assembly. 
In this handout, we introduce some commonly used methods in phylogenetic community ecology.
These methods will allow us to account for phylogenetic diversity and provide insight into the mechansisms that give rise to the over- and under-dispersion (i.e., clustering) of biological communities. 

After completing this exercise you will know how to:

1.  measure phylogenetic $\alpha$ diversity
2.  measure phylogenetic $\beta$ diversity
3.  evaluate the contribution of phylogeny to geographical patterns of biodiversity

## Directions:
1. Change "Student Name" on line 3 (above) with your name.
2. Complete as much of the exercise as possible during class; what you do not complete in class will need to be done on your own outside of class.
3. Use the handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
4. Be sure to **answer the questions** in this exercise document; they also correspond to the handout.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. For homework, follow the directions at the bottom of this file. 
7. When you are done, **Knit** the text and code into a PDF file.
8. After Knitting, please submit the completed exercise by creating a **pull request** via GitHub.
Your pull request should include this file *PhyloTraits_exercise.Rmd* and the PDF output of `Knitr` (*PhyloTraits_exercise.pdf*).

## 1) SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  
1. clear your R environment,  
2. print your current working directory,  
3. set your working directory to your "*/PhyloCom*" folder, 
4. load all of the required R packages (be sure to install if needed), and  
5. load the required R source file.

```{r}










```

## 2) DESCRIPTION OF DATA
We will revisit the data that was used in the Geographical Ecology module. 
As a reminder, in 2013 we sampled ~ 50 forested ponds located in Brown County State Park, Yellowood State Park, and Hoosier National Forest in southern Indiana. 
In addition to measuring a suite of geographic and environmental variables, we characterized the diversity of bacteria in the ponds using molecular-based approaches.
Specifically, we amplified the 16S rRNA gene (i.e., "DNA") and 16S rRNA transcripts (i.e., "RNA") of bacteria using barcoded primers on the Illumina MiSeq platform.
We then used a `mothur` pipeline to quality-trim our data set and assign sequences to operational taxonomic units (OTU).
In this exercise, we will use the DNA sequence data for making community phylogenetic inference.  

## 3) LOADING OF DATA
In the R code chunk below, do the following:  
1. load the environmental data for the Brown County ponds (*20130801_PondDataMod.csv*),  
2. load the site-by-species matrix using the `read.otu()` function,  
3. subset the data to include only the DNA,  
4. rename to remove extra charactors in site names,  
5. remove any sites or OTUs in the site-by-species that need to be removed, and   
7. load the taxanomic data using the `read.tax()` function.

```{r}







```

In the R code chunk below, do the following:  
1. load the fasta alignment for the OTUs,  
2. rename the OTUs be removing everything before the tab (\t) and after the bar (|),  
3. import the Methanosarcina outgroup fasta file,  
4. convert both fasta files into the DNAbin format and combine using `rbind()`,  
5. visualize the sequence alignment,  
6. using the alignment (with outgroup), create a distance matrix (use any evolutionary model),  
7. using the distance matrix above, make a neighbor joining tree
8. remove any tips (OTUs) that are not in the community data set
9. plot the rooted tree

```{r}









```

## 4) PHYLOGENETIC ALPHA DIVERSITY

### A.  Faith's Phylogenetic Diversity (PD)
In the R code chunk below, do the following:  
1. calculate Faith's D using the `pd()` function.

```{r}


```

In the R code chunk below, do the following:  
1. make a biplot for species richness and Faith's D,  
2. add the trend line, and  
3. calculate the scaling exponent. 

```{r}





```

***Question 1***: Answer the following questions about the PD-SR pattern.    
a.  Describe the relationship between taxonomic richness and phylodiversity?  
b.  Mathematically, why should they be correlated?  
c.  Under what conditions would you expect these two estimates of diversity to deviate from one another?  

> ***Answer 1a***:  
> ***Answer 1b***:  
> ***Answer 1c***:  

#### i.  Randomizations and Null Models

In the R code chunk below, do the following:
1. estimate the standardized effect size of PD using the richness randomization method, and  
2. pick two additional randomization methods and repeat. 

```{r}




```

***Question 2***: Using `help()` and the table above, identify two null models that can be used with the `ses.pd()` function.
Run `ses.pd()` with these null models and answer the following questions:

a.  What hypotheses are being tested with the p-values associated with `ses.pd`?
b.  What features might affect the intepretation of the `ses.pd` output?

> ***Answer 2a***:  
> ***Answer 2b***:  

### B.  Phylogenetic Dispersion Within a Sample
Another way to assess phylogenetic $\alpha$-diversity is to look at dispersion within a sample. 

#### i. Phylogenetic Resemblance Matrix 

In the R code chunk below, do the following:
1. calculat the phylogenetic resemblance matrix.

```{r}


```

#### ii.  Net Relatedness Index (NRI)
In the R code chunk below, do the following:
1. Calculate the NRI for each site in the Brown County ponds data set. 

```{r}






```

### iii.  Nearest Taxon Index (NTI)  
In the R code chunk below, do the following:
1. Calculate the NTI for each site in the Brown County ponds data set. 

```{r}



```

***Question 3***: In the NRI and NTI examples above, the arguments "abundance.weighted = FALSE" means that the indices were calculated using presence-absence data.
Modify and rerun the code so that NRI and NTI are caculated using abundance data. 
How does this affect the interpretation of NRI and NTI?

> ***Answer 3***:  

### C) Mini-Exercise On Phylogenetic Alpha-Diversity
Conduct an exploratory analysis of the phylogenetic $\alpha$-diversity for the Indiana refuge ponds.
Using techniques from the *Week1_Handout.Rmd* and any other statistical methods at your disposal, identify enviornmental variables that may influence the taxonomic richness (SR), phylodiveristy (PD), and dispersion (NRI and NTI) of the bacterial communities in the sampled ponds. 
Generate some hypotheses as to what ecological and/or evolutionary processes are giving rise to these patterns of phylogenetic $\alpha$-diversity.  

```{r}



```

> ***Mini-Exercise: Code, Output, and Discussion***:


## 5) PHYLOGENETIC BETA DIVERSITY 

### A. Phylogenetically Based Community Resemblance Matrix
As you may recall, in order to quantify $\beta$-diversity for more than two samples, one needs to create a resemblance matrix. 
When quantifying taxonomic $\beta$-diversity, we calculated the pairwise **similarity** or **dissimiliarity** for all samples in a site-by-species matrix using metrics such as the Sørensen index or the Bray-Curtis index. 
More or less, we need to go through the same process for quantifying phylogenetic $\beta$-diversity.
Instead of making the resemblance matrix based on incidence or abundance of taxa, we are going to incorporate information about the phylogenetic relationships among taxa. 
Similar to other measures of $\beta$-diversity, there are numerous ways to calculate the phylogenetic distances in the community resemblance matrix.
Here, we will explore two: Mean Pairwise Distance and UniFrac distance.

\begin{center}
\hyphenpenalty 10000
\exhyphenpenalty 10000
\begin{tabular}{ m{5cm}  m{10cm} }
  \textbf{Index} & \textbf{Description} \\
  \hline \hline \\ [-1.5ex]
  \textbf{Mean Pairwise Distance} & 
  Distance between two samples calculated as the mean phylogenetic distance between pairs of taxa \\
  \\ [-1.5ex]
  \textbf{UniFrac} & 
  Distance between two samples calculated as $\Sigma_{unshared}$ / $\Sigma_{total}$, where $\Sigma_{unshared}$ is the sum of unshared branch lengths between samples and $\Sigma_{total}$ is the total (shared and unshared) branch lengths in a rooted tree \\
  \\ [-1.5ex]
  \hline
\end{tabular}
\end{center}

Let's calculate the phylogenetically based community resemblance matrices for our pond data set.

In the R code chunk below, do the following:
1. calculate the phylogenetically based community resemblance matrix using mean pairwise distance, and  
2. calculate the phylogenetically based community resemblance matrix using UniFrac distance.

```{r}




```

In the R code chunk below, do the following:
1. plot mean pair distance versus UniFrac distance and compare.

```{r}





```

***Question 4***: Using the plot above, describe the relationship between Mean Pair Distance and UniFrac distance. 
Note: we are calcuating unweighted phylogenetic distances (similar to incidence based measures).
That means that we are not taking into account the abundance of each taxa in each site.

> ***Answer 4***: 


### B. Visualizing Phylogenetic Beta Diversity
Now that we have our phylogenetically based community resemblance matrix, we can visualize phyogenetic diversity among samples using the same techniques that we used in the $\beta$-diversity module.

In the R code chunk below, do the following:  
1. perform a PCoA based on the UniFrac distances, and  
2. calculat the explained variation for the first three axes.

```{r}






```

Now that we have calculated our PCoA, we can plot the results.

In the R code chunk below, do the following:  
1. plot the PCoA results, 
2. include the appropriat axes,  
3. be sure to add and label the points, and  
4. customize the plot. 

```{r}












```

***Question 5***: How much variation is explained in the first three PCoA axes if Mean Pair Distance is used as the resemblance matrix instead of UniFrac?

> ***Answer 5***: 

### C. Hypothesis Testing

#### i. Categorical Approach: Watershed Effect  
The ponds that we sampled in southern Indiana were located in three distinct watersheds (BCSP, HNF, and YSF). 
For many organisms, watershed boundaries represent major dispersal barriers, which may influence the phylogenetic distribution of species. 
Here, we will test for this watershed affect using the Permutational Multivariate Analysis of Variance (PERMANOVA) test that we learned about in the $\beta$-diversity module. 

```{r, results = 'hide'}
# Define Environmental Category
water.shed <- env$Location

# Run PERMANOVA with `adonis()` Function {vegan}
adonis(dist.uf ~ water.shed, permutations = 999)
```

#### ii. Continuous Approach: Environmental Gradients  
In the Geographic Ecology module, we demonstrated that there was substantial variation in enviornmental variables that are known to influence the structure and function of microbial communities.
In the following section, we will revist two methods that are used to test for relationships among multivariate environmental and biological data. 

First, let's define the environmental data and then create an enviornmental distance matrix:

```{r}
# Define Environmental Variables
envs <- env[, 5:19]

# Distance Matrix for Environmental Variables
env.dist <- vegdist(scale(envs), method = "euclid")
```

Second, let's conduct a Mantel's test to see whether UniFrac distance is correlated with environmental variation:

```{r, results = 'hide'}
# Conduct Mantel Test {vegan}
mantel(dist.uf, env.dist) 
```

Last, let's conduct a Canonical Correspondence Analysis (CCA). 
You will recall that this constrained ordination technique allows one to test for the effects of an explanatory matrix (e.g., environmental data) on a response matrix (e.g., phylogenetic distance matrix). 

```{r, results = 'hide'}
# Conduct CCA {vegan}
ponds.cca <- vegan::cca(dist.uf ~ scale(envs))

# Permutation Tests: Axes and Env Variables
anova(ponds.cca, by = "axis")
ponds.fit <- envfit(ponds.cca, envs, perm = 999)
ponds.fit

# Calculate Explained Variation
cca.explainvar1 <- round(ponds.cca$CCA$eig[1] / 
                         sum(c(ponds.cca$CCA$eig, ponds.cca$CA$eig)), 3) * 100
cca.explainvar2 <- round(ponds.cca$CCA$eig[2] / 
                         sum(c(ponds.cca$CCA$eig, ponds.cca$CA$eig)), 3) * 100

# Define Plot Parameters
par(mar = c(5, 5, 4, 4) + 0.1)

# Initiate Plot
plot(scores(ponds.cca, display = "wa"), xlim = c(-3, 3), ylim = c(-3.5, 3),
     xlab = paste("CCA 1 (", cca.explainvar1, "%)", sep = ""),
     ylab = paste("CCA 2 (", cca.explainvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1.2, axes = FALSE)

# Add Axes
axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

# Add Points & Labels
points(scores(ponds.cca, display = "wa"),
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(scores(ponds.cca, display = "wa"), 
    labels = row.names(scores(ponds.cca, display = "wa")), cex = 0.5)

# Add Environmental Vectors
vectors <- scores(ponds.cca, display = "bp")
row.names(vectors) <- c("Elev", "Dia", "Dep", "Vol", "ORP", "Temp", "SpC", "DO", 
                        "TDS", "Sal", "pH", "Col", "Chla", "DOC", "DON")
arrows(0, 0, vectors[,1] * 2, vectors[, 2] * 2, 
       lwd = 2, lty = 1, length = 0.2, col = "red")
text(vectors[,1] * 2, vectors[, 2] * 2, pos = 3, 
     labels = row.names(vectors))
axis(side = 3, lwd.ticks = 2, cex.axis = 1.2, las = 1, col = "red", lwd = 2.2,
     at = pretty(range(vectors[, 1])) * 2, labels = pretty(range(vectors[, 1])))
axis(side = 4, lwd.ticks = 2, cex.axis = 1.2, las = 1, col = "red", lwd = 2.2,
     at = pretty(range(vectors[, 2])) * 2, labels = pretty(range(vectors[, 2])))
```

***Question 6***: Based on the multivariate procecures conducted above, describe some of the phylogenetic patterns of $\beta$-diversity for bacterial communities in the Indiana refuge ponds.   

> ***Answer 6***: 

### D) Mini-Exercise On Phylogenetic Beta-Diversity
Generate and test hypotheses about the phylogenetic $\beta$-diversity of the Indiana refuge ponds.
Using the environemtanal variables identified in the $\alpha$-diversity mini-exercise, redo either the PERMANOVA or CCA analysis above using just these variables. 
Feel free to add any additional approaches we used in *Beta_Handout.Rmd*. 
Compare your results to the findings from above and discuss any differences. 

```{r}





```

> ***Mini-Exercise Discussion***:


## 6) GEOGRAPHICAL PHYLOGENETIC COMMUNITY ECOLOGY

### A. Phylogenetic Distance-Decay (PDD)
You will recall from the Geographical Ecology module, that the distance decay (DD) relationship reflects the spatial autocorrelation of community similarity.
That is, geographically near communities should be more similar than geographically distant communities.
Here, we will test to what degree spatial autocorrelation can also affect phylogenetic DD.  

First, we will calcualte distances for geographic data, taxonomic data, and phylogenetic data among all unique pair-wise combinations of ponds.

```{r}
# Geographic Distances (Kilometers) Among Ponds
long.lat <- as.matrix(cbind(env$long, env$lat))
coord.dist <- earth.dist(long.lat, dist = TRUE)

# Taxonomic Distances Among Ponds (Bray-Curits)
bray.curtis.dist <- 1 - vegdist(comm) 

# Phylogenetic Distances Among Ponds
unifrac.dist <- 1 - dist.uf 
  
# Transform All Distances Into List Format:
unifrac.dist.ls <- liste(unifrac.dist, entry = "unifrac")
bray.curtis.dist.ls <- liste(bray.curtis.dist, entry = "bray.curtis")
coord.dist.ls <- liste(coord.dist, entry = "geo.dist")

# Create a Data Frame from the Lists of Distances
df <- data.frame(coord.dist.ls, bray.curtis.dist.ls[, 3], unifrac.dist.ls[, 3])
names(df)[4:5] <- c("bray.curtis", "unifrac")
attach(df)
```

Now, let's plot the DD relationships:

```{r, results = 'hide'}
# Set Initial Plot Parameters
par(mfrow=c(2, 1), mar = c(1, 5, 2, 1) + 0.1, oma = c(2, 0, 0, 0))

# Make Plot for Taxonomic DD
plot(coord.dist, bray.curtis, xlab = "", xaxt = "n", las = 1, ylim = c(0.1, 0.9),
     ylab="Bray-Curtis Similarity", 
     main = "Distance Decay", col = "SteelBlue")

# Regression for Taxonomic DD 
DD.reg.bc <- lm(bray.curtis ~ geo.dist)
summary(DD.reg.bc)
abline(DD.reg.bc , col = "red4")

# New Plot Parameters
par(mar = c(2, 5, 1, 1) + 0.1)

# Make Plot for Phylogenetic DD
plot(coord.dist, unifrac, xlab = "", las = 1, ylim = c(0.1, 0.9),
     ylab = "Unifrac Similarity", col = "darkorchid4")

# Regression for Phylogenetic DD
DD.reg.uni <- lm(unifrac ~ coord.dist)
summary(DD.reg.uni)
abline(DD.reg.uni, col = "red4")

# Add X-Axis Label to Plot
mtext("Geographic Distance, km", side = 1, adj = 0.55, 
      line = 0.5, outer = TRUE)
```

Finally, let's test whether the slopes for taxonomic and phylogenetic DD are significantly different from one another.

```{r, results = 'hide'}
diffslope(geo.dist, unifrac, geo.dist, bray.curtis)
```

***Question 7***: Interpret the slopes from the taxonomic and phylogenetic DD relationships. If there are differences, hypothesize why this might be.   

> ***Answer 7***: 


### A. Phlogenetic diversity-area relationship (PDAR)

Helmus and Ives (2012) developed methods to study how phylogenetic diversity changes with increasing area.
This phylogenetic diversity-area relationship (PDAR) is analogous to the species-area relationship (SAR) that we learned about in the Geographical Ecology module.
You will recall that the SAR is a cumulative relationship (S = cA^z) and so, cannot be negative.
In fact, the shallowest the SAR can get is to have a slope of 0.0, meaning that all species are found in all samples.
Below is a plot of the taxonomically based SAR for the pond data. 

```{r, echo=FALSE, results = 'hide'}
SAR <- function(OTUs){
  Alist <- c()
  Slist <- c()
  
  num.ponds <- c(2,4,8,16,32,51)
  for (i in num.ponds) {   
    areas <- c() # hold iterated area values 
    Ss <- c() # hold iterated S values
    
    for(j in 1:100){
      pond.sample <- sample(51, replace = FALSE, size = i) 
      area <- 0
      cum.abs <- vector(mode = "numeric", length = length(comm))
  
      for (k in pond.sample) { # Loop through each randomly drawn pond
        area <- area + pond.areas[k] # aggregating area, (incrementing)
        cum.abs <- cum.abs + as.numeric(comm[k, ])
    
      } # End random pond samples loop
      S <- length(cum.abs[cum.abs > 0])
      areas <- c(areas, area)
      Ss <- c(Ss, S)
    }
    Alist <- rbind(Alist, mean(areas))
    Slist <- rbind(Slist, mean(Ss))
    print(c(mean(areas), mean(Ss)))
  }
  return(cbind(log10(Alist), log10(Slist)))
}

pond.areas <- as.vector(pi * (env$Diameter/2)^2) # Find areas of all 51 ponds
sar <- SAR(comm)
sar <- as.data.frame(sar)

plot(sar, xlab = "log(Area)", ylab = "log(Richness)", 
     main = "Species-Area Relationship", col = "red")

OLS <- lm(sar$V2 ~ sar$V1)
abline(OLS, col = "red", lw = 2)

slope <- round(coefficients(OLS)[2], 3)
legend("topleft", legend = paste("slope =", slope), bty = "n", lw = 2, col = "red")  
```

In contrast to SAR, phylogenetic divesity-area relationships can increase or decrease with area. 
This provoked Helmus and Ives (2012) to develop theoretical expectations for the effects that ecological and evolutionary processes might have on the shape of the PDAR.

#### i. Constructing the PDAR  
Helmus and Ives (2012) used the phylogenetic species variability (PSV) metric to quantify phylogenetic diversity.
PSV quantifies how phylogenetic relatedness decreases the variance of a hypothetical neutral trait shared by all species in a community. 
Type 'help(psv)' to learn more.

In this exercise, we will construct PDARs using the random aggregation approach of Helmus and Ives (2012).
This approach is similar to how we constructed SARs in the Geographical Ecology module.

Let's begin by writing a function to generate the PDAR.

```{r}
PDAR <- function(comm, tree){

  # Create Objects to Hold Areas and Diversity
  areas <- c() 
  diversity <- c() 
  
  # Create Vector Increasing Number of Plots by 2x
  num.plots <- c(2, 4, 8, 16, 32, 51)
  
  for (i in num.plots){   
    # Create vectors to hold areas and diversity form iterations, used for means
    areas.iter <- c() 
    diversity.iter <- c() 
    
    # Iterate 10 Times Per Sample Size
    for (j in 1:10){ 
      # Sample w/o replacement
      pond.sample <- sample(51, replace = FALSE, size = i)
      
      # Create Variable and Vector to Hold Accumulating Area and Taxa
      area <- 0 
      sites <- c() 
      
      for (k in pond.sample) {       # Loop Through Each Randomly Drawn Pond  
        area <- area + pond.areas[k] # Aggregating Area (Roughly Doubling)
        sites <- rbind(sites, comm[k, ])    # And Sites
        }
      
      # Concatenate the rea to areas.iter
      areas.iter <- c(areas.iter, area)  
      # Calculate PSV or Other Phylogenetic Alpha-Diversity Metric
      psv.vals <- psv(sites, tree, compute.var = FALSE) 
      psv <- psv.vals$PSVs[1] 
      diversity.iter <- c(diversity.iter, as.numeric(psv))
      }

    diversity <- c(diversity, mean(diversity.iter)) # Let Diversity be the Mean PSV
    areas <- c(areas, mean(areas.iter))             # Let areas be the Average Area
    print(c(i, mean(diversity.iter), mean(areas.iter))) # Print As We Go
    }
  # Return Vectors of Areas (x) and Diversity (y)
  return(cbind(areas, diversity)) 
  }
```

#### ii. Evaluating the PDAR  
We will examine the relationship between phylogenetic diversity and area using both Spearman's correlation coefficient (S) and Pearson's correlation coefficient (P).
It is informative to use both because while S is computed on ranks and depicts monotonic relationships (the degree to which the relationship is continually increasing or decreasing), P is computed on the observed values and therefore depicts linear relationships.

```{r, results = 'hide'}
# Calculate Areas for Ponds: Find Areas of All 51 ponds
pond.areas <- as.vector(pi * (env$Diameter/2)^2) 

# Compute the PDAR
pdar <- PDAR(comm, phy) 
pdar <- as.data.frame(pdar)
pdar$areas <- sqrt(pdar$areas)

# Calculate Perason's Correlation Coefficient
Pearson <- cor.test(pdar$areas, pdar$diversity, method = "pearson")
P <- round(Pearson$estimate, 2)
Pp <- round(Pearson$p.value, 3)

# Calculate Spearman's Correlation Coefficient
Spearman <- cor.test(pdar$areas, pdar$diversity, method = "spearman")
S <- round(Spearman$estimate, 2)
Sp <- round(Spearman$p.value, 3)

# Plot the PDAR
par(mar = c(5, 5, 4, 2) + 0.1)
plot(pdar[, 1], pdar[, 2], xlab = "Area", ylab = "PSV", ylim = c(0, 1), 
     main = "Phylogenetic Diversity-Area Relationship", 
     col = "red", pch = 16, las = 1)

legend("topleft", legend= c(paste("Spearman Correlation = ", S, "; p = ", Sp, sep = ""), 
                            paste("Pearson Correlation = ", P, "; p = ", Pp, sep = "")), 
                            bty = "n", col = "red") 
```

***Question 8***: For the bacteria in the Indiana ponds, the slope (*z*) of the SAR was 0.14.
This is slightly lower than *z*-values observed for many macroscopic organisms (e.g., fish, birds, plants), but is higher than what has been reported for other microbial systems. 
However, what did we observe for the microbial PDAR in the Indiana pond?
How might we explain the differences between the taxonomic (SAR) and phylogenetic (PDAR)?  

> ***Answer 8***: 


## 7) HOMEWORK 

1.  In their study of phylogenetic diversity-area relationships (PDARs), Helmus and Ives, increased area by aggregating plots in two ways.
First, they aggregated plots with respect to whether the plots were adjacent; what they called 'spatial'.
Second, they aggregated plots at random ('non-spatial').
While both spatial and non-spatial sampling methods capture the effect of increasing area, sampling with respect to location (i.e., spatial) also captures the effect of increasing distance or spatial autocorrelation.
Explain the general difference that Helmus and Ives (2012) observed between spatial and non-spatial PDARs, and why this difference should be expected in both the species-area relationship (SAR) and the phylogenetic diversity-area relationship.

2.  Use Knitr to create a pdf of your completed PhyloTraits_handout.Rmd document, push it to GitHub, and create a pull request.
The due date for this assignment is March 4, 2015 at 12:00 PM (noon).
