---
title: "Phylogenetic Diversity - Traits"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "February 26, 2015"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW
Taxonomic measures of $\alpha$- and $\beta$-diversity do not incorporate the evolutionary information into estimates of communities. 
In this handout, we will introduce methods used in phylogenetic community ecology.
These methods will allow us to account for phylogenetic diversity and provide insight into the mechansisms that give rise to over- and under-dispersion (i.e., clustering) of biological communities. 

After completing this exercise you will know how to:

1.  measure phylogenetic $\alpha$ diversity
2.  measure phylogenetic $\beta$ diversity
3.  construct phylogenetic diversity-area relationships

## 1) SETUP
### A. Retrieve and Set Your Working Directory

```{r, results = 'hide'}
rm(list = ls())
getwd()
setwd("~/GitHub/QuantitativeBiodiversity/Assignments/PhyloCom")
```

### B. Load Packages 
We wil be relying heavily on the R package `picante`. 
This package has many of the functions that are contained in the software Phylocom, which is used for the analysis of phylogenetic community structure and character evolution (http://phylodiversity.net/phylocom/).

After the initial installation of these packages using the `install.packages()` fucntion, let's load the packages and their dependencies with the `require()` function:

```{r, results = 'hide', message = FALSE, warning = FALSE} 
require("picante")
```

## 2) DESCRIPTION OF DATA
We will revisit the environmental, geographic, and biological data that we used a couple of weeks ago tht was collected from the Brown County refuge ponds. 

## 3) LOADING OF DATA
```{r}
# Examples from Picante Vignette (http://goo.gl/EYOZ55)
data(phylocom)
names(phylocom)

# Phylogentic Data: Tree of Class "Phylo"
# Phylo object requires {ape}
# mothur Generates a Newick formatted file so we'll probably need to use the `read.tree` function in {ape}
# We need to make sure that the names associate with the tree matach up with the names in the site-by-species matrix in `comm` below. 

phy <-(phylocom$phy)

# Community Data: Site-by-Species Matrix
comm <- phylocom$sample

# Traits Data: I don't think we'll using this, and least not for ponds
traits <- phylocom$traits
```

```{r}
# Just working through Kembel example
# This is actually plotting samples onto  the tree; labels just correspond to "mock" samples, which demonstrate different patterns of dispersion

prunedphy <- prune.sample(comm, phy)

par(mfrow = c(2, 3))

for (i in row.names(comm)) {
  plot(prunedphy, show.tip.label = FALSE, main = i) 
  tiplabels(tip = which(prunedphy$tip.label %in% names(which(comm [i, ] > 0))), pch = 19, cex = 2)
  }
```

## 4. PHYLOGENETIC ALPHA DIVERSITY
"One of the earliest measures of phylogenetic relatedness in ecological communities
was the phylogenetic diversity (PD) index proposed by Faith (1992). 
Faith's PD is
definned as the total branch length spanned by the tree including all species in a local
community. 
The `pd` function returns two values for each community, the PD and the
species richness (SR)"

```{r}
# compares PD and taxonomic richness (SR)
pd.result <- pd(comm, phy, include.root=FALSE)
pd.result
```


```{r}
# plot phylo vs. taxo diversity
par(mar = c(5, 5, 2, 1) + 0.1)
plot(pd.result$SR, pd.result$PD, 
     pch = 20, col = "red", las = 1, asp = 1, xlim = c(0, 35), ylim = c(0, 35),
     xlab = "Taxonomic Richnes", ylab = "Phylodiversity") 
abline(b = 1, a = 0, lty = 2)
text(20, 30, "1:1")

# correlation betwen phylo and taxo
corr <- cor(pd.result$SR, pd.result$PD)
```

***Question 2***: Compare patterns of phylogenetic diversity and species richness. 
a.  Does taxonomic richness and phylodiversity correlate?
b.  Mathematically, why should they be correlated?  
c.  Under what conditions would you expect these two estimates of diversity to deiate from one another?

## MPD, MNTD, SES(MPD),SES(MNTD)

```{r}
phydist <- cophenetic(phy)
ses.mpd.result <- ses.mpd(comm, phydist, null.model="taxa.labels", abundance.weighted = FALSE, runs = 99)
ses.mpd.result

ses.mntd.result <- ses.mntd(comm, phydist, null.model="taxa.labels", abundance.weighted = FASLE, rungs = 99)
ses.mntd.result
```

## PHYLOGENETIC BETA DIVERSITY 

```{r}
comdist.result <- comdist(comm, phydist)
comdist.result
comdist.clusters <- hclust(comdist.result)
plot(comdist.clusters)
```

Ordinations: PCoA taxo  vs. UNIFRAC
http://www.inside-r.org/packages/cran/picante/docs/unifrac
http://cran.r-project.org/web/packages/GUniFrac/GUniFrac.pdf
http://joey711.github.io/phyloseq-demo/unifrac.html

## 7) HOMEWORK

1.  Below is the output of a multiple regression model depicting the relationship between the maximum growth rate ($\mu_{max}$) of each bacterial isolate and the niche breadth of that isoalte on the 18 different sources of phosphorus. 
One feature of the study which we did not reveal earlier in the handout is that the isolates came from two different lakes. 
One of the lakes is an very oligotrophic (i.e., low phosphorus) ecosystem named Little Long (LL) Lake. 
The other lake is an extremely eutrophic (i.e., high phosphorus) ecosystem named Wintergreen (WG) Lake.
We included a "dummy variable" (D) in the multiple regression model (0 = WG, 1 = LL) to account for the environment from which the bacteria were obtained.  
Based on your knowledge of the traits and their phylogenetic distributions, what conclusions would you draw about our data and the evidence for a generalist-specialist tradeoff? 

2.  Use Knitr to create a pdf of your completed PhyloTraits_handout.Rmd document, push it to GitHub, and create a pull request.
The due date for this assignment is February 25, 2015 at 12:00 PM (noon).
