---
title: "Phylogenetic Diversity - Traits"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "February 20, 2015"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW
Up to this point, we have been focusing on patterns taxonomic diversity in Quantitative Biodiversity. 
Taxonomic diversity is certainly an important dimension of biodiversity, but it fails to consider the evolutionary history and relatedness of species. 
The goal of this exercise is to introduce basic concepts of phylogenetic diversity. 

After completing this exercise you will know how to:

1.  create phylogenetic trees
2.  map traits onto phylogenetic trees
3.  test for phylogenetic signal of traits on a phylogenetic tree
4.  test for relationships among traits using phylogenetic independent contrasts 

## 1) SETUP
### A. Retrieve and Set Your Working Directory

```{r, results = 'hide'}
rm(list = ls())
getwd()
setwd("~/GitHub/QuantitativeBiodiversity/Assignments/PhyloTraits")
```

### B. Load Packages 
There are many contributed R packages that have been developed for conducting phylogenetic analyses (http://cran.r-project.org/web/views/Phylogenetics.html). 
We will rely heavily on `ape`, which peforms a wide range of phylogenetic analyses. 
We will supplement this with other packages for some specialized tasks such as sequence alignment (`muscle`), reading of alinged sequences (`seqinr`), and maximum likelihood reconstruction (`phangorn`).
Let's install these packages using the `install.packages()` fucntion. 
Then let's load the packages and their dependencies with the `require()` function:

```{r, results = 'hide', message = FALSE, warning = FALSE} 
require("muscle")
require("ape")
require("seqinr")
require("phangorn")
require("geiger")
```

## 2) DESCRIPTION OF DATA
The maintenance of biodiversity is thought to be influenced by trade-offs among species. 
One such trade-off involves the ability of a highly specialized species to peform well on a single type of resource. 
This specialization may be advantageous under some conditions, but disadvantageous if the species finds itself in a habitat where resource conditions change.
In contrast, some species are capable of using many different resources, but it is assumed that there may be a fitness cost associated with being a generalist. 
We set out to test this hypothesized trade-off in bacteria by assessing generalist and specialist strategies on different forms of phopshorus, which commonly limits growth and reproduction of organisims inhabiting freshwater ecosystems.
We collected 40 strains of bacteria from two lakes. 
After identifying these bacteria based on their 16S rRNA gene sequences, we measured their growth on 18 different forms of phosphorus. 
We will take a phylogenetic approach to mapping phosphorus resource use onto a phylogenetic tree while testing for specialist-generalist trade-offs. 

## 3) SEQUENCE ALIGNMENT
After generating or retrieving sequence data, the first step of a phylogenetic analysis is to peform an alignment. 
This can be done with nucleotides (DNA or RNA) or proteins (amino acids). 
There are various methods for aligning sequences, but the basic premise is to use a algorithm to arrange sequences relative to one another using conserved regions so that evolutionary reconstructions can be performed.
In R, there are at least two packages that can be used for aligning sequences: `ape` and `muscle`. 
In `ape`, alignments can be done with the functions `clustal`, `muscle`, and `tcoffee`, all which have a range of options. 
However, these functions require that other programs be installed on your computer.
Therefore, for this exericse, we are going to use the `muscle` package in R becasue it will align sequences in a "stand alone" fashion without the need for installing additional software. 

```{r}
seq.aln <- muscle::muscle(seqs="./data/p.isolates.fasta", out = "./data/p.isolates.afa") 
```

## 4) VISUALIZING ALIGNMENTS
It is good practice to view your sequences after alignment. 
This may not be practical for large databases, but in our example, we're only dealing with 45 sequences plus an outgroup sequence. 
First, we will read in the fasta-aligned file that we created above using the `read.alignment` function from the `seqinr` package. 
Then, using the `ape` package, we will convert the alignnment file it into a **DNAbin object**, which is a bit-level coding scheme that allows R to store and manipulate sequence data.
Last, we will visualize the alignment by color-coding bases different colors. 

```{r}
read.aln <- read.alignment(file = "./data/p.isolates.afa", format = "fasta") 
p.DNAbin <- as.DNAbin(read.aln) 
x <- p.DNAbin[,100:500] # identify base pair region of 16S rRNa gene to visuzlize
image(x, cex.lab = 0.5) # ape command to visusalize seq align
grid(ncol(x),nrow(x),col="lightgrey") # ape command to add grid
```

***Question 1***:  Make some observations about the `muscle` alingmnet of the 16S rRNA gene sequences for our bacterial isoaltes. Approximatley how long were our reads? 

> ***Answer 1***: 

## 5) MODELS OF DNA EVOLUTION
Phylogenetic analyses require assumptions about the rates of DNA evolution. 
Various models have been developed to estimate the rates that nucleotides change, which can be important for inferring relatedness of genes.
For example, the **Jukes-Cantor model (JC69)** is the simplest **substitution model**.
It assumes all that nucleotides occur at equal frequencies and that these nucleotides can mutate from one to another with equal probability. 
The **Felenstein model (F81)** builds from JC69 by allowing base frequences to vary. 
The **Kimura model (K80)** also assumes equal frequencies of nucleotides, but recognizes that transition mutations (e.g., purine to a purine [A -> G] or pyrmidine to pyrmidine [C -> T]) occur with higher probability than tranversion mutations (e.g., purine to pyrmidine or vice-versa).
**Hasegawa, Kishino, and Yano (HKY85)** merges elements of F81 and K80 by accounting for uneuqal base frequencies and differences in the rates of transitions and tranversions. 
Many other models of DNA evolution exist, which take into account features such as G+C bias and the probabilities associated with time-reversible nucleotide changes (e.g., T -> C -> T).
Fortunately, the `ape` package has a function (`dist.dna`) that will create distances for a large number of DNA substitution models. 
Let's create a distance file based on the Kimura (K80) subsitution model using the `dist.dna` function in `ape`. 

```{r}
seq.dist <- dist.dna(p.DNAbin, model = "K80", pairwise.deletion = FALSE) # 
```

## 6) PHYLOGENETIC RECONSTRUCTION
Phylogenetic trees...

### NEIGHBOR JOINING
Neighbor joining trees take a distance matrix based on comparisons of pairs of sequences.
Trees are then constructed in an agglomerative fashion similar to the methods that we used in cluster analysis in the $\beta$-diversity exercise.
Neighbor joining is a relatively fast algorithm and often generates correct tree topologies. 
However, today, other non-distance-based phylogenetic methods are more commonly used. 

***NOT SURE WE NEED TO DO BOTH NJ AND UPGMA***
```{r}
nj.tree <- nj(seq.dist) # ape command for Neighbor Joining
outgroup <- match("Methanosarcina", nj.tree$tip.label)
nj.rooted <- root(nj.tree, outgroup, resolve.root=TRUE)
#upgma.tree <- upgma(seq.dist)
#upgma.rooted <- root(upgma.tree, outgroup, resolve.root=TRUE)
#layout(matrix(c(1,2), 2, 1), height=c(2,2))
#par(mar = c(1,1,1,1))
plot.phylo(nj.rooted, main = "Neigbor Joining Tree", "phylogram", use.edge.length = FALSE, direction = "right", cex = 0.6, label.offset = 1) # ape command
#plot(upgma.rooted, main = "UPGMA", use.edge.length = FALSE, direction = "right", cex = 0.5, label.offset = 1)
```

## 7) MAXIMIM PARSIMONY
Using package `phangorn`.
Requires phyDat formatted objects.
There are functions like `optim.parsimony` that peforms tree rearragements via nearest-neighbor exchange and `pratchet` which uses parsimony ratchet to find trees with lower parsimony score.
Type the following at the R terminal for a detailed exmaple: vignette('Trees')

```{r}
# read.aln.mat <- as.matrix(read.aln)
# read.aln.phydat <- phyDat(read.aln.mat, type = "DNA")
# seq.dist.phydat <- dist.dna(as.DNAbin(read.aln.phydat))
# tree.nj <- nj(seq.dist.phydat)
# parsimony(tree.nj, read.aln.phydat)
# tree.pars <- optim.parsimony(tree.nj, read.aln.phydat)
# tree.ratch <- pratchet(read.aln.phydat, trace = 0)
# parsimony(c(tree.pars,tree.ratch),read.aln.phydat)
# plot(tree.pars, main = "parsimony", use.edge.length = FALSE, direction = "right", cex = 0.5, label.offset = 1)
```

## 6) MAXIMUM LIKELIHOOD ESTIMATION AND BOOTSTRAP SUPPORT
** Beware really long, something is not right. 
** Discussed by Paradis (2012) on pp 138-159
** Problems with `pml`
### `myphylo` has been removed!!! Advice is to use phangorn
** We should consider `mlphylo` in `ape`
ml <- mlphylo(p.DNAbin, model = DNAmodel("JC69")) 

```{r}
read.aln.mat <- as.matrix(read.aln)
read.aln.phydat <- phyDat(read.aln.mat, type = "DNA")
tree.p <- NJ(dist.logDet(read.aln.phydat))
ml <- pml(tree.p, read.aln.phydat, k = 4, inv = 0.2)
fit.opt <- optim.pml(ml, optGamma = TRUE, optInv=TRUE)

# data(yeast) # yeast is a list of `phyDat` tpe
# tree <- NJ(dist.logDet(yeast))
# fit <- pml(tree, yeast, k = 4, inv = 0.2)
# fit.opt <- optim.pml(fit, optGamma = TRUE, optInv=TRUE)

```

## 7) MAPPING TRAITS ONTO TREES

### A. Loading Trait Database
```{r}
p.growth <- read.table("./data/p.isolates.stdgrowth.txt", sep = "\t", header = TRUE, row.names = 1)
#p.growth <- as.data.frame(t(p.growth))
```

### Mapping Traits on Tree
```{r}
nj.tree <- bionj(seq.dist) # ape command for Neighbor Joining
outgroup <- match("Methanosarcina", nj.tree$tip.label)
nj.rooted <- root(nj.tree, outgroup, resolve.root=TRUE)
nj.rooted <- drop.tip(nj.rooted, "Methanosarcina")
traits <- p.growth[,c("ATP", "SRP", "PolyP", "Phyt", "AEP")]

# Drop Missing Branches and data
nj.rooted <- drop.tip(nj.rooted, c("LL11A", "LL15", "WG37", "WG30A", "WG32", "LL12A"))
traits <- traits[!rownames(traits) == "LL12", ]

require(phylobase)
require(adephylo)
x <- phylo4d(nj.rooted, traits)
table.phylo4d(x,treetype="clad", show.node=FALSE, cex.leg=0.5, cex.label = 0.5,
use.edge.length=FALSE, edge.color="blue", edge.width=2, box=FALSE, las = 1) 




library(latticeExtra)
dd.col <- as.dendrogram(hc)
col.ord <- order.dendrogram(dd.col)
mat <- matrix(rnorm(4*4),nrow=4)
colnames(mat) <- tree$tip.label
rownames(mat) <- tree$tip.label
levelplot(mat[tree$tip,tree$tip],type=c('g','p'),
          aspect = "fill",
          colorkey = list(space = "left"),
          legend =
            list(right =
                   list(fun = dendrogramGrob,
                        args =
                          list(x = dd.col, 
                               side = "right",
                               size = 10))),
          panel=function(...){
            panel.fill('black',alpha=0.2)
            panel.levelplot.points(...,cex=12,pch=23)
          }
)




```


### B. Calculating Niche Breadth
We calcualte the niche breadth (*nb*) index from Levins (1968) which is defined as $\frac{1}{\left( n \cdot \Sigma p_{xi}^{2}}\right)$, where ni is each resource 

single isolate
```{r}
levins <- function(p_xi = ""){
  x = 0
  for (i in p_xi){
    x = x + i^2
    }
    nb = 1 / (length(p_xi) * x) 
  return(nb)
}
```  

```{r}
levins.2 <- function(p_xi = ""){
  p_xi = as.numeric(p_xi)
  nb = 1 / (length(p_xi) * (p_xi %*% p_xi))
  return(nb)
}  
levins.2(p.growth[1, ])

```

## 8) PHYLOGENETIC SIGNAL
Phylogenetic signal occurs when the traits of closely related species are more similar than what is expected by chance. 
In many cases, testing for phylogenetic signal is a prerequsite for addressing questions related to trait evolution and phylogenetic biodiversity. 
If there is significant phylogenetic signal, then evolutionary history is important for understanding the relationships between for example a trait and the peformance of individuals being considered. 
If there is *not* any phylogenetic signal, then the relationship between traits and peformance can explored using normal statistical methods. 

### Pagel’s Lambda
http://bodegaphylo.wikispot.org/iv._testing_phylogenetic_signal_in_r
http://blog.phytools.org/2013/10/simplified-lambdatree.html


```{r}
require("geiger")
lambdaTree(read.aln.phydat, 0) ---> needs to be class phylo!!!

```




## 9) PHYLOGENETIC INDEPENDENT CONTRASTS
**Phylogenetic independent contrasts are outlined in Paradis (2012) on pp 204-209
**Can we take a trait-peformance relationship that's siginficant and correct for phylogeny? 
Is the relationship still significant? 
Do the parameters or explained varition change?

