---
title: "Phylogenetic Diversity - Traits"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "February 20, 2015"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

## 1) SETUP
### A. Retrieve and Set Your Working Directory

```{r, results = 'hide'}
rm(list = ls())
getwd()
setwd("~/GitHub/QuantitativeBiodiversity/Assignments/PhyloTraits")
```

### B. Load Packages 
```{r, results = 'hide', message = FALSE, warning = FALSE} 
require("muscle")
require("ape")
require("seqinr")
```

## 2) DESCRIPTION OF DATA
The meintenance of biodiversity is thought to be influenced by trade-offs among species. 
One such trade-off involves species that are highly specialized at using a single resource. 
However, this may come at the cost of not being able to use other resoruces as well.
We set out to test this hypothesized trade-off in bacteria by assessing generalist and specialist strategies on different forms of phopshorus, which commonly limits growth and reproduction of populations in freshwater ecosystems.
More..

## 3) SEQUENCE ALIGNMENT
It appears that DNA sequences can be alligned in R using at least two packages: `ape` and `muscle`. 
In `ape`, alignments can be done with the functions `clustal`, `muscle`, and `tcoffee`, all which seem to have a range of options. 
However, it seem that these functions call on other installed programs, like ClustalX. 
In contrast, the `muscle` package aligns sequences in a "stand alone" fashion. 

```{r}
seq.aln <- muscle::muscle(seqs="p.isolates.fasta") # align seqs; more options?!?!
muscle::write.fasta(seq.aln, file = "out.aln.muscle") # save muscle-alligned file
#seqinr::write.fasta(sequences=seq.aln[[2]], seq.aln[[1]], file.out="p.isolates.aln.fasta")
```

## 4) VISUALIZE THE ALIGNMENT
```{r}
#read.aln <- read.alignment(file = "p.isolates.aln.fasta", format = "fasta")
read.aln.m <- read.alignment(file = "out.aln.muscle", format = "fasta") # seqinr command to read muscle align
p.DNAbin <- as.DNAbin(read.aln.m) # ape command to convert fasta file into DNAbin filetype
x <- p.DNAbin[,400:500] # identify region (in bp of 16S rRNA) to visuzlize
image(x, cex.lab = 0.5) # ape command to visusalize seq align
grid(ncol(x),nrow(x),col="lightgrey") # ape command to add grid
```

## 5) CREATE EVOLUTIONARY MODEL AND NEIGHBOR-JOINING TREE
```{r}
seq.dist <- dist.dna(p.DNAbin, model = "K80", pairwise.deletion = TRUE)
nj.tree <- nj(seq.dist)
```




