---
title: "Phylogenetic Diversity - Traits"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "February 20, 2015"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## 1) SETUP
### A. Retrieve and Set Your Working Directory

```{r, results = 'hide'}
rm(list = ls())
getwd()
setwd("~/GitHub/QuantitativeBiodiversity/Assignments/PhyloTraits")
require("heR.Misc")

### A. Loading Growth Rate Data
```{r}
p.growth <- read.table("data/p.isolates.raw.growth.txt", sep = "\t", header = TRUE, row.names = 1)
#p.growth <- as.data.frame(t(p.growth))
```

### B. Calculate Nich Breadth
We wil estimate niche breadth using Levins index (1968).
Levins index ranges from 0 - 1 with lower values indicating specialization and higher values indicating generalization. 
To make the growth rate data mesh with the indes, we need to standardize by taking the sum of the growth rates of an isoalte across all resoruces. 

```{r}
p.growth.std <- p.growth / (apply(p.growth, 1, sum))
```

Make function for Levins Index
```{r}
levins <- function(p_xi = ""){
  x = 0
  for (i in p_xi){
    x = x + i^2
    }
    nb = 1 / (length(p_xi) * x) 
  return(nb)
}
```  

Niche breadth for isolates on different P sources
```{r}
nb <- levins(p.growth.std)
```

### C. Testing for Generalist-Specialist Tradeoff
We will test for a hypothesized trade-off between niche breadth and a strain's maximum growth rate. 
The expectation is that generalists will experience a cost that will be manifested in an overall lower maximum growth rate. 

The following code gets the data set up for testing
```{r}
umax <- (apply(p.growth, 1, max)) # calculate max growth
vec <- rep(NA, length(umax)) # make an empty vector for lake id
tradeoff <- data.frame(nb,umax,vec) # make new data frame
lake.test <- grepl("WG",row.names(tradeoff)) # test if WG or LL
lake <- ifelse(lake.name == TRUE, 'WG', 'LL') # create lake id vector
tradeoff <- data.frame(nb, umax, lake) # make dataframe
tradeoff.LL <- tradeoff[ which(tradeoff$lake=='LL'),] ## LL data
tradeoff.WG <- tradeoff[ which(tradeoff$lake=='WG'),]
```

The following dummy variable regression tests for trade-off
Specifically, the model test for an effect of niche breadth (continuous), lake identity (categorical), and an interaction between the two variables
```{r}
D <- (lake=="LL")*1
fit<-lm(log10(umax) ~ nb + D + nb*D)
summary(fit)
```

### Plotting
LL.nb <- subset(nb,lake=="LL")
WG.nb <- subset(nb,lake=="WG")

LL.umax <- subset(umax,lake=="LL")
WG.umax<-subset(umax,lake=="WG")

plot(LL.nb, LL.umax, axes = F, xlab = "Niche Breadth", ylab = "Maximum Growth Rate", 
  log = "y", pch = 21, cex = 2, las = 1, col = "black", bg = "white", 
  xlim=c(0,1), ylim = c(min(umax-0.5), max(umax+0.5)))


plot(LL.nb, LL.umax, xlab = "Niche Breadth", ylab = "Maximum Growth Rate", 
  log = "y",pch = 21, cex = 2, col = "black", bg = "white", 
  xlim=c(0,1), ylim = c(min(umax-0.5), max(umax+0.5)), las = 2)

points(WG.nb,WG.umax,pch=21,cex=2,col="black", bg="black")

axis(3,labels=F)
axis(4,labels=F)

plot(LL.nb,LL.umax,xlab="niche breadth",ylab="Maximum Growth Rate", pch=21, cex=2, col="black", bg="white",xlim=c(0,1),ylim=c(1.25*(min(umax)),1.25*(max(umax))), las = 2)

axis(3,labels=F)
axis(4,labels=F)

legend(0.005, 0.7,"LL", cex=1, col="black", pch=1, bty="n")
legend(0.005, 0.4,"WG", cex=1, col="black", pch=16, bty="n")

### Model Coefficients
coefficients<-coefficients(fit) # model coefficients
confit<-confint(fit,level=0.95) # 95% CI for parameters
predicted<-fitted(fit) # predicted values
rediduals<-residuals(fit) # residuals
anova<-anova(fit) # anova table
vcov<-vcov(fit) # covariance matrix for model parameters
influence<-influence(fit) # regression diagnostics

### Adding Regression Lines
curve(coefficients[1]+coefficients[2]*x,from=min(nb),to=max(nb),add=TRUE)
Int=coefficients[1]+coefficients[3]
Slp=coefficients[2]+coefficients[4]
curve((Int)+(Slp)*x,from=min(nb),to=max(nb),add=TRUE)
