---
title: "Week 1 Exercise: Basic R"
author: "Student Name; Z620: Quantitative Biodiversity, Indiana University"
date: "January 16, 2015"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise introduces some of the basic features of the R (http://www.r-project.org/) computing environment. 
We will briefly cover operators, data types, and simple commands that will be useful for you during the course and beyond. 
In addition to using R's base package, we will also use contributed packages, which together will allow us to visualize data and peform relatively simple statistics (e.g., linear regression and ANOVA).

## Directions:
Change **Student Name** above (line 3) with your name.
You are working in an RMarkdown file.
It will allow you to integrate text and R code into one file.
There are two major feature to this document: 1) Markdown formatted text 2) R code chunks.
Anything in an R code chunk will be intrepreted by R when you *Knit* the document.
Please complete as much of this as possible during class; what you do not complete in class will need to be done on your own outside of class.
Use the handout (hardcopy) as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
Be sure to answer the questions in the handout in this document.
Space for your answer is provided and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is all highlighted by RStudio in green.
Before you leave the classroom today, it is *imperative* that you push this file to your GitHub repo.
For homework, follow the directions at the bottom of this file. 
When you are done, **Knit** the text and code into a PDF file.
Basically, just press the `Knit` button in the scripting panel above.
After Knitting, please submit by creating a **pull request** via GitHub. 
(Review your inclass Git assignment if needed.)
The completed exercise is due on **January 21^st^, 2015 before 12:00 PM (noon)**.

## 1) HOW WE WILL BE USING R AND OTHER TOOLS

During the course, we will use RStudio (http://www.rstudio.com/), which is a user-friendly integrated developement environment (IDE) that allows R to interface with other tools. 
For example, the document you are reading was generated in RMarkdown (<http://rmarkdown.rstudio.com>). 
Markdown is a simple formatting syntax for authoring HTML, PDF, and other documents. 

We will also use a tool called knitr (http://yihui.name/knitr/), which is a package that generates reports from R script and Markdown text. 
For example, when you click the **Knit PDF** button in the scripting window of RStudio, a document will be generated that includes LaTeX (http://www.latex-project.org/) typesetting as well as the output of any embedded R code. 

However, if there are errors in the Rcode contained in your Markdown document, you will not be able to knit a PDF file. 
Assignments in this class will require that you successfully create a Markdown-generated PDF using knitr; you will then **push** this document to the course **respository** hosted on IU's GitHub (<https://github.iu.edu>) and generate a **pull request**.

## 2) SETTING YOUR WORKING DIRECTORY

In the R code chunk below, please provide the code to: 
1) clear your R environment, 
2) print your current working directory, and 
3) set your working directory to your Week1 folder.
```{r}



```

## 3) USING R AS A CALCULATOR

In the R code chunk below, please provide examples of each: addition, subtraction, multiplication, and division.
```{r}




```

In the R code chunk below, please calculate the following: 
1) the volumn of a cube with length, l, = 5, 
2) the area of a circle with radius,r, = 2 (area = pi * r^2),  
3) the length of the opposite side of a right-triangle given that the angle, theta, = pi/4 (radians, AKA 45°) and with hypotenuse lenght sqrt(2) (remember: sin(theta) = opposite/hypotenuse), 
4) the log (base e) of your favorite number.

```{r}




```

## 4) ASSIGNING VARIABLES

In the R code chunk below, do the following:
1) Create an object "a" and assign any number to it.
2) Create a second object "b" and set it equal to 2 x "a".
3) Show the value of "b".
4) Next, reassign to a different number to "a".
5) Finally, show "b"" again to demonstrate that R did not update "b" when "a" was changed (i.e., is not a symbolic programming language).
6) When you are done, clear your R working environment using the `rm()` function.

```{r}






```

## 5) WORKING WITH VECTORS

### Basic Features Of Vectors

In the R code chunk below, do the following: 
1) Create a vector x consisting of any 5 numbers.
2) Create a vector w that is 5 times x.
3) Add x and w and divide by 6

```{r}



```

Now, in the R code chunk below, do the following: 
1) Create another vector (k) that is the same length as w.
2) Multiply k by x (do not use matrix multiplication).
3) Combine w and k into a third vector that consists of any three elements from w and any four elements of k.

```{r}



```

### Summary Statictics of Vectors

In the R code chunk below, calcualte the summary statistics (maximum, minimum, sum, mean, median, variance, standard deviation, and standard error of the mean) for the vector (v) provided.

```{r}
v <- c(16.4, 16.0, 10.1, 16.8, 20.5, NA, 20.2, 13.1, 24.8, 20.2, 25.0, 20.5, 30.5, 31.4, 27.1)













```

## 5) WORKING WITH MATRICES

In the R code chunk below, do the following:
Using a mixture of Approach 1 and 2, create a matrix with two columns and 5 rows.
Both columns should consist of random numbers.
Make the mean of the first 8 and the mean of the second 25.
In addition, adjust for the first column set sd = 2 and for the second column set sd = 10.

```{r}





```

***Question 1***: What does the rnorm function do? What are the arguments specifying?

> Answer 1:


In the R code chunk below, do the following: 
1) Load `matrix.txt` from the Week1 data folder as matrix "m"
2) Transpose this matrix
3) Determine the dimensions of the tranposed matrix

```{r}



```


***Question 2***: What are the dimensions of the matrix you just transposed?

> Answer 2:


###Indexing a Matrix

In the R code chunk below, do the following:
1) Index matrix m selecting all but the third column
2) Remove the last row of matrix m

```{r}


```

***Question 3***: Describe what we just did in the last indexing operation in the handout

> ***Answer 3***:


## 6) BASIC DATA VISUALIZATION AND STATISTICAL ANALYSIS
### Load Zooplankton Dataset

In the R code chunk below, do the following:
1) Load the zooplankton dataset from the Week1 data folder.
2) Display the structure of this data set.

```{r}


```

### Correlation

In the R code chunk below, do the following:
1) Create a matrix with the numerical data in the 'meso' dataframe.
2) Visualize the pairwise **bi-plots** of the 6 numerical variables.
3) Conduct a simple **Pearson's correlation** analysis

```{r}


```


***Question 4***: Describe some of the general features based on the visualization and correlation anaylysis above?

> Answer 4:


In the R code chunk below, do the following:
1) Redo the correlation analysis using the `corr.test()` function in the "psych" package with the following options: method = "pearson", adjust = "BH".
2) Now, redo this correlation analysis using a non-parametric method.
3) Use the print command from the handout to see the results of each.

```{r}



```

***Question 5***: Describe what you learned from corr.test. 
Specifically, are the results sensitive to whether you use parametric (i.e., Pearson's) or non-parametric methods?
With the Pearson's method, is there evidence for false discovery rate due to multiple comparisons?

> ***Answer 5***:

In the R code chunk below, use the `corrplot` function in the "corrplot" package to produce the ellipse correlatoin plot seen in the handout.

```{r}


```

### Linear Regression

In the R code chunk below, do the following:
1) Conduct a linear regression analysis to test the relationship between total nitrogen (TN) and zooplankton biomass (ZP)
2) Examine the output of the regression analaysis
3) Produce a plot of this regression analysis including the following: categorically labeled points, the predicted regressionline with 95% confidence intervals, and the appropriate axis labels.

```{r}







```

*** Question 6***: Interpret the results from the regression model

> ***Answer 6***:



*** Question 7***: Explain what the `predict()` function

> ***Answer 7***:

Using the R code chunk below, use the code provided in the handout to determine if our data meet the assumptions of the linear regression analysis. 

```{r}

```

+ Upper left: is there a random distribtion of the residuals around zero (horizontal line)?
+ Upper right: is there a resonably linear relationship between standardized residuals and theoretical quantiles? Try `help(qqplot)`
+ Bottom left: again, looking for a random distribution of sqrt(standardized residuals)
+ Borrom right: leverage indicates the influence of points; contours correspond with Cook's distance, where values > |1| are "suspicious"

### ANALYSIS OF VARIANCE (ANOVA)

Using the R code chunk below, do the following:
1) Order the nutrient treatments from low to high
2) Produce a barplot to visualize zooplankton biomass in each nutrient treatment
3) Be sure to include error bars on your plot and label the axis appropriately.
4) Use a one-way analysis of variance to determine if the treatments differed with respond to zooplankton biomass
5) Use Tukey's HSD to conduct a post-hoc comparision

```{r}







```

***Question 8***: How do you interpret the ANOVA results relative to the regression results?
Do you have any concerns about this analysis?

> ***Answer 8***:


Using the R code chunk below, use the code provided in the handout to determine if our data meet the assumptions of the analysis of variance. 

```{r}

```

## HOMEWORK

1) Complete the R code chunks above.
1) Complete this entire exercise (Week1_Exercise.Rm, in week 1 folder)

2) In addition, redo the section on linear regression (including summary statistics, plotting, and diagnostics) using log10-transformed zooplankton biomas data. Does the transformation help meet assumption of linear regression or change the interpretation of the analysis in anyway? Describe.

3) Use Knitr to create a pdf of your completed Week1_Exercise.Rmd document, push it to GitHub, and create a pull request.
This assignment is due on **January 21^st^, 2015 at 12:00 PM (noon)**.
