---
title: "Week 1 Exercise: R Test"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "January 16, 2015"
output: pdf_document
geometry: margin=2.54cm
---

The goal of this short assignment to to make sure that your personal computing environment is functioning as needed for the course.
From the RMarkdown file (`R_test.Rmd`), you will print a simple statement and then retrieve information about the operating system and versions of software on your computer.
You will then  **Knit** the text and code into a PDF file.
Last, you will submit this as your first assignment, which is due on **January 20^th^, 2015 at 12:00 PM**

1. Edit your name in the *author* line above (line 3)

2. Print a message in the R terminal (***do you mean the "Console Pane?") using the `print` command:

    Ex. Print "*Hello, World!*"
```{r}
print("Hello, World!")
```

    Now, print your own statement (it can be anything):
```{r}


```

3. Now we'll run the code below to retrieve information about your local computing system. 
You do not need to make any changes to the code. 
The output below is what was returned from Mario's computer:

```{r}
info <- unname(c(Sys.info()[6],
          .Platform$OS.type,
          Sys.info()[1:2],
          R.version.string,
          .Platform$GUI,
          system("git --version", intern = TRUE)))
info
```

