---
title: "Week 2 Exercise"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "November 8, 2014"
output: pdf_document
---
In this exercise, we will conduct exercises on alpha diversity

##RETRIEVE AND SET YOUR WORKING DIRECTORY##
```{r}
getwd()
setwd("/Users/lennonj/GitHub/Quantitative_Biodiversity/Assignments/Week2")
```

##INSTALL PACKAGES##
People develop different packages for certain tasks that can be carried out in the R enviornment. Use the 'help'funciton to learn about package installation and add-ons. install.packages("vegan")
```{r}
require("vegan")
library("vegan")
```
In the library of vegan, there is a data set that we will be using called BCI. BCI stands for Barro Colorado Island, which is a located in Panama. The BCI data frame has 50 plots (rows) of 1 hectare with counts of trees on each plot with total of 225 species (columns)
```{r}
data(BCI)
```
Let's look at the data in the first few plots
```{r}
head(BCI)
```
Calculate Shannon index. 'Margin = 1' means diversity is calculated row-wise; 'Margin = 2' means diversity is calcualted column-wise.With base = exp(1), we are esimating Shannon's index using the natural logrithm of each taxon's relative abundance using the equation; H` = -Si pi ln (Pi)
```{r}
shannon<-diversity(BCI,index="shannon", MARGIN=1,base=exp(1))
```
Can also calculate other diversity metrics
```{r}
simpson <- diversity(BCI, "simpson")
invsimpson <- diversity(BCI, "inv")
fisher <- fisher.alpha(BCI)
```
Let's plot
```{r}
pairs(cbind(shannon, simpson, invsimpson, fisher), pch="+", col="blue")
```
Species richness
```{r}
S <- specnumber(BCI)
```
Rarefaction
```{r}
(raremax <- min(rowSums(BCI)))
Srare <- rarefy(BCI, raremax)
plot(S, Srare, xlab = "Observed No. of Species", ylab = "Rarefied No. of Species")
abline(0, 1)
rarecurve(BCI, step = 20, sample = raremax, col = "blue", cex = 0.6)
```

http://www.jennajacobs.org/R/rarefaction.html
