---
title: "Among Site (Beta) Diversity - Supplementary"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "February 6, 2015"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---
## OVERVIEW
In this exercise, you will explore between-site biodiversity, also known as beta ($\beta$) diversity. 
First, you will learn how to formally quantify $\beta$-diversity.
Second, you will learn how to visualize $\beta$-diversity using cluster analysis, ordination, and heatmaps. 
Last, you will be introduced to some multivariate techniques to test hypotheses about $\beta$-diversity. 

## 1) SETUP
### Retrieve and Set Your Working Directory

```{r, results = 'hide'}
rm(list=ls())
getwd()
setwd("~/GitHub/QuantitativeBiodiversity/Assignments/Beta")
```

### Load Packages 

From last week's exercises, you should now be familiar with some of the features in the R package `vegan`.
We will be using this pacakge again; please load it now.

```{r, results = 'hide', message = FALSE, warning = FALSE} 
require("vegan")
```

## 2) LOADING DATA
We will again be using the BCI data set contained in `vegan` for part of this week's exercise. 
In addition, we will be using environmental data for the BCI plots. 
Your Beta folder has soil data for each of the BCI plots (bci.soil.txt).
Let's go ahead and import all of these data now.

```{r}
data(BCI)                                   # BCI Tree Abundance (vegan)
BCI.soil <- read.delim("./bci.soil.txt")    # BCI Soil Data (your working directory)
```

## 3) Exploring $\beta$-Diversity
Last week, we learned about $\alpha$-diversity, where we calculated things like species richness and evenness. 
These metrics are great for assessing the diversity of a single sample, but overlook spatial and temporal heterogeneity.
Beta $\beta$-diversity is a measure of between-habitat diversity, or the difference in species composition between two or more samples. 