---
title: "Among Site (Beta) Diversity - Supplementary"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "February 6, 2015"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

We will again be using the BCI data set contained in `vegan` for part of this week's exercise. 
In addition, we will be using environmental data for the BCI plots. 
Your Beta folder has soil data for each of the BCI plots (bci.soil.txt).
Let's go ahead and import all of these data now.

```{r}
data(BCI)                                   
BCI.soil <- read.delim("./data/bci.soil.txt")    
```

## 3) Exploring $\beta$-Diversity
Last week, we learned about $\alpha$-diversity, where we calculated things like species richness and evenness. 
These metrics are great for assessing the diversity of a single sample, but overlook spatial and temporal heterogeneity.
$\beta$-diversity is a measure of between-habitat diversity, or the difference in species composition between two or more samples. 

### BCI Species Richness
To illustrate the importance of $\beta$-diversity, let's go back to the BCI dataset and examine tree richness among all sites. 
First, let's calculate richness for the BCI sites:

```{r}
BCI.S <- rowSums(BCI > 0) * 1
```

Now, we are going to visualize tree species richness by ploting sites based on their locations.
We are going to color-code the sites based on tree species richness using a **heatmap**.
In R there are a few tools that allow you to make color pallets. 
Here we are going to use `terrain.colors` but you can learn about other colors families by looking at the `Pallettes` help file: `help(Pallettes)`.
Let's start by making a color pallette for Richness (S):

```{r}
BCI.S.color <- rev(terrain.colors(max(BCI.S) + 1))
```

Notice that we reversed ('rev()') the sequence of colors in the pallette, this was done to make the color ranges more intuitive (darker colors correspond with larger values of S). 
Also, notice that we have to define the number of colors in each pallette. 
Here we used one more than the max values for S. 
This is done to deal with any 0s in our data.

Let's start by plotting tree richness for each plot at BCI. 
But how do we know how these sites are arranged? 
Well lucky for us, the BCI soil dataset includes standardized XY-coordinates for each plot:

```{r}
head(BCI.soil, n = 2L)
```

Here we are going to plot sites in a standardized XY-coordinate space and use our heatmap color to display tree species richness for each of the 50 sites.
In addition, we will add a legend to the plot so that readers understand the heatmap scale.
This plot also includes a bunch of new commands that are important for producing effective figures.
To learn more about these commands, you can modify them or use the `help()` files. Take a few minutes to investigate.

```{r, fig.width=6, fig.height=3.5}
# BCI Tree Richness
par(mar=c(4,4,3,4) + 0.1, xpd=TRUE) # xpd allows the legend to appear outside of the plot
plot(BCI.soil$x, BCI.soil$y, asp = 1, pch = 22, cex = 4, las = 1, 
     bg = BCI.S.color[BCI.S + 1], 
     xlim = c(0, 1000), ylim = c(0, 500), main = "BCI Richness (S)", 
     xlab = "X-Coordinate (m)", ylab = "Y-Coordinate (m)")
text(BCI.soil$x, BCI.soil$y, BCI.S, cex = 0.5, col = "red")
legend("topright", inset = c(-0.2, 0), pch = 22, pt.cex = 3, title = "S", 
       bty ='n', legend = seq(0, 150, 25), pt.bg = BCI.S.color[seq(1, 151, 25)])
```

**Questions**

A. What do you notice when you compare all 50 sites? 

```
    Please Provide Answer Here
```

B. Is this a good way to compare the sites?

```
    Please Provide Answer Here
```

C. What does this method of comparing sites miss?

```
    Please Provide Answer Here
```

### BCI Individual Tree Abunance
Patterns in richness may obscure differences in tree species composition across the BCI landscape.
To investigate this, let's just focus on the abundance of a specific tree: the Prioria tree. 

First, let's generate a heatmap pallette for tree abundance.
We will plot the abundance data in the same way we did for richness.

```{r}
BCI.N.color <- rev(terrain.colors(41))
```


```{r, fig.width=6, fig.height=3.5}
# BCI Prioria Abundance
par(mar=c(4,4,3,4) + 0.1, xpd=TRUE)
plot(BCI.soil$x, BCI.soil$y, asp = 1, pch = 22, cex = 4, las = 1, 
     bg = BCI.N.color[BCI$Prioria.copaifera + 1], 
     xlim = c(0,1000), ylim = c(0, 500), main = "Prioria Abundance (N)", 
     xlab = "X-Coordinate (m)", ylab = "Y-Coordinate (m)")
text(BCI.soil$x, BCI.soil$y, BCI$Prioria.copaifera, cex = 0.5, col="red")
legend("topright", inset=c(-0.2, 0), pch=22, pt.cex=3, bty='n',
       title="N", legend=seq(0, 40, 10), pt.bg = BCI.N.color[seq(1, 41, 10)])
```

**Questions**

D. What do you observe across BCI?

```
    Please Provide Answer Here
```

E. How does this compare to the patterns of species richness?

```
    Please Provide Answer Here
```
Notice that the doubs data set is actully a list with 4 components (elements).
Lists might be new data structures for you. 
In R, a list is an ordered collection of objects. 
The first component of the doubs list is the environmental data for each of the 30 sites (doubs$env). 
There are 11 environmental variables in this dataset. 
See the help file for more information on each (including units). 
The second component are the abundances at each site for 27 fish species. 
The third component has the xy spatial coordinates for each site. 
The last component contains the names of each fish species. 

Let's plot fish richness in these stream communities.
We will first need to calculate richness and define our color pallettes.
Once we have these items we can plot stream fish richness in a similar fashion as we did with BCI tree communities.
Below are three plots similar to the BCI plots we created above.
These plots show richness of fish at all sites in the Doubs River.

```{r, fig.width=5, fig.height=4.5, echo=TRUE}
# Stream Fish
par(mar = c(4, 4, 3, 5) + 0.1, xpd = TRUE)    
spa.S <- specnumber(doubs$fish)               
spa.S.color <- rev(terrain.colors(31))    # Define Richness Color Pallette
spa.N.color <- rev(terrain.colors(7))     # Define Abundance Color Pallette

# Stream Fish Richness
plot(doubs$xy, asp = 1, type = 'l', col = "light blue", lwd = 10, las = 1, 
     xlim = c(0,280), ylim = c(0,280), main = "Fish Richneses (S)", 
     xlab = "X-Coordinate (km)", ylab = "Y-Coordinate (km)")
points(doubs$xy, pch = 22, cex=2, bg = spa.S.color[spa.S + 1])
text(doubs$xy, as.character(spa.S), cex = 0.5, col="red")
text(150, 0, "Upstream", cex = 1, col = "red")
text(48, 114, "Downstream", cex = 1, col = "red")
legend("topright", inset=c(-0.25, 0), pch = 22, pt.cex = 2, bty = 'n',
       title = "S", legend = seq(0, 30, 10), pt.bg = spa.S.color[seq(1, 31, 10)])
```

Notice that the code we used here is almost identical to that we used with the BCI plots.
Though creating a single figure in R may seem like a lot of work, it is beneficial when making multiple plots because you can recycle code easily.
However, sometimes there are slight variations (e.g., the blue line in the stream plot).

Below are the plots for two individual fish species in the Doubs River.
You should be able to code this with ease using the Stream Fish Richness plot code and the BCI abundance plot code. 
You can try this on your own time, but here we are just going to provide the plots for you.

```{r, fig.width=8, fig.height=3.75, echo=FALSE}
# Define Plot Parameters
par(mfrow = c(1,2), mar = c(4, 4, 3, 4) + 0.1, xpd = TRUE)    

# Brown Trout Abundance
plot(doubs$xy, asp = 1, type = 'l', col = "light blue", lwd = 10, las = 1, cex.axis = 0.75, 
     xlim = c(0, 280), ylim = c(0, 280),main = "Brown Trout Abundance (N)", 
     xlab = "X-Coordinate (km)", ylab = "Y-Coordinate (km)")
points(doubs$xy, pch = 22, cex=2, bg = spa.N.color[doubs$fish$Satr + 1])
text(doubs$xy, as.character(doubs$fish$Satr), cex = 0.5, col="red")
text(150, 0, "Upstream", cex=1, col="red")
text(48, 114, "Downstream", cex=1, col="red")
legend("topright", inset=c(-0.25, 0), pch = 22, pt.cex = 2, bty = 'n',
       title = "N", legend = seq(0, 6, 2), pt.bg = spa.N.color[seq(1, 7, 2)])

# Common Roach Abundance
plot(doubs$xy, asp = 1, type = 'l', col = "light blue", lwd = 10, las = 1, cex.axis = 0.75, 
     xlim = c(0, 280), ylim = c(0, 280),main = "Common Roach Abundance (N)", 
     xlab = "X-Coordinate (km)", ylab = "Y-Coordinate (km)")
points(doubs$xy, pch = 22, cex=2, bg = spa.N.color[doubs$fish$Ruru + 1])
text(doubs$xy, as.character(doubs$fish$Ruru), cex = 0.5, col="red")
text(150, 0, "Upstream", cex = 1, col = "red")
text(48, 114, "Downstream", cex = 1, col = "red")
legend("topright", inset=c(-0.25, 0), pch = 22, pt.cex = 2, bty = 'n',
       title = "N", legend = seq(0, 6, 2), pt.bg = spa.N.color[seq(1, 7, 2)])
```

F. How does this dataset differ from the BCI data? 

```
    Please Provide Answer Here
```

G. Is richness the most appropriate tool to compare communities, why or why not?

```
    Please Provide Answer Here
```

H. What about other $\alpha$-diversity measures?

```
    Please Provide Answer Here
```

This week we are going to compare the diversity of communities across sites. 
We will start by comparing diversity across sites. 






### NMDS
```{r}
spe.nmds <- metaMDS(spe, distance="bray")
spe.nmds
spe.nmds$stress
plot(spe.nmds, type="t", main=paste("nMDS/Bray - Stress =", round(spe.nmds$stress, 3)))
abline(h=0, lty=3)
abline(v=0, lty=3)
```

What is the nMDS stress? How is this used to judge the quality of the ordination





