---
title: "Week 3 Exercise"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "January 23, 2015"
output: pdf_document
---
In this exercise, we will conduct exercises on beta diversity

## RETRIEVE AND SET YOUR WORKING DIRECTORY

```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/QuantitativeBiodiversity/Assignments/Week2")
```

## INSTALL PACKAGES
People develop different packages for certain tasks that can be carried out in 
the R enviornment. 
Use the 'help'funciton to learn about package installation and add-ons.

```{r}
require("vegan")||install.packages("vegan");require("vegan")
require("ade4")||install.packages("ade4");require("ade4")
require("BiodiversityR")||install.packages("BiodiversityR");require("BiodiversityR")

```

In the library of vegan, there is a data set that we will be using called BCI. 
BCI stands for Barro Colorado Island, which is a located in Panama. 
The BCI data frame has 50 plots (rows) of 1 hectare with counts of trees on 
each plot with total of 225 species (columns).

We will again be using the BCI data set for part of this weeks exercises

```{r}
data(BCI)
data(BCI.env)
BCI.soil <- read.delim ('http://www.davidzeleny.net/anadat-r/lib/exe/fetch.php?media=data:bci.soil.txt')
data(doubs)
```


## Recap
S <- specnumber(BCI)
simpson <- diversity(BCI, "simpson")
invsimpson <- diversity(BCI, "inv") 

spa <- doubs$xy

windows()
par(mar=c(5,5,4,6) + 0.1, xpd=TRUE)
plot(spa, asp=1, cex=doubs$fish$Satr, main="Brown Trout", xlab = "X-Coordinate (km)", 
ylab = "Y-Coordinate (km)")
lines(spa, col="light blue")
text(spa, row.names(spa), cex-0.8, col="red")
text(45, 10, "Upstream", cex=1.2, col="red")
text(38, 115, "Downstream", cex=1.2, col="red")
legend("topright", inset=c(-0.25, 0), legend=c(1,2,3,4,5), pch=1, pt.cex=c(1,2,3,4,5), title="Fish", bty='n')


## Turnover




## Visualizaton


### Cluter Analysis


### Ordinations


## Hypothesis Testing




## Homework
